---
title: "clustering"
author: "Will Jones"
date: "April 24, 2016"
output: 
  html_document: 
    keep_md: yes
    toc: yes
---

Our goal here is to find appropriate ways to cluster the stations by their
patterns bikes in and bikes out through the day. 

```{r packages, echo = FALSE, message=FALSE, warning=FALSE, results='hide'}
packages <- c("dplyr", "ggplot2", "data.table", "lubridate", "ggmap")
sapply(packages, library, character.only = TRUE)
```


```{r load-data}
load("stations.RData")
stations$id = as.factor(stations$id)
trips <- fread("data/hubway_trips.csv", header = T, sep = ',')
trips <- trips %>%
  mutate(status = factor(status),
         start_date = mdy_hms(start_date),
         end_date = mdy_hms(end_date),
         bike_nr = factor(bike_nr),
         subsc_type = factor(subsc_type),
         gender = factor(gender),
         zip_code = factor(zip_code),
         strt_statn = factor(strt_statn),
         end_statn = factor(end_statn))

# Create a start time of day variable
trips <- mutate(trips, start_time = start_date)
day(trips$start_time) <- 1
month(trips$start_time) <- 1
year(trips$start_time) <- 2015
trips <- trips %>%
  mutate(start_time = as.numeric(start_time),
         start_time = start_time - min(start_time),
         start_time = start_time / 3600) # convert time from seconds to hours

# Create an end time of day variable
trips <- mutate(trips, end_time = end_date)
day(trips$end_time) <- 1
month(trips$end_time) <- 1
year(trips$end_time) <- 2015
trips <- trips %>%
  mutate(end_time = as.numeric(end_time),
         end_time = end_time - min(end_time),
         end_time = end_time / 3600) # convert time from seconds to hours
```


```{r aggregate-rides}
trips %>% group_by(strt_statn) %>% summarise(count = n()) %>% arrange(desc(count))
```

Morning rush hour approx time: 7--10am
Evening rush hour approx time: 4--7pm

Features:

- proportion of bikes out during morning rush hour
- proportion of bikes in during morning rush hour
- proportion of bikes out during evening rush hour
- proportion of bikes in during evening rush hour
- variance in hourly rates

## Choosing Features

```{r feature-extraction}
bike_out <- trips %>%
  group_by(strt_statn) %>%
  summarise(prop_out_morning = sum(between(start_time, 7, 10)) / n(),
            prop_out_evening = sum(between(start_time, 16, 19)) / n())

bike_in <- trips %>%
  group_by(end_statn) %>%
  summarise(prop_in_morning = sum(between(end_time, 7, 10)) / n(),
            prop_in_evening = sum(between(end_time, 16, 19)) / n())


# Merge features
stations <- left_join(stations, bike_out, by = c("id" = "strt_statn")) %>%
  left_join(bike_in, by = c("id" = "end_statn"))
```

## Clustering stations

```{r clustering}
clustering_data <- select(stations, prop_out_morning, prop_out_evening, prop_in_morning, prop_in_evening)

test_k <- function(k) kmeans(clustering_data,k, nstart = 40)$tot.withinss
ss <- sapply(1:8, test_k)

qplot(x = 1:8, y = ss) + geom_line() +
  labs(title = "Total Within SS by Choice of k",
       x = "k",
       y = "Total Within SS") + theme_bw()

clustering <-  kmeans(clustering_data, centers = 3, nstart = 40)
pc <- prcomp(clustering_data)

stations$cluster <- as.factor(clustering$cluster)
stations <- cbind(stations, pc$x) %>% data.frame() %>% tbl_df()

```

```{r results}
qplot(data = stations, x = prop_out_morning, y = prop_out_evening, color = cluster)
qplot(data = stations, x = prop_in_morning, y = prop_in_evening, color = cluster)
qplot(data = stations, x = PC1, y = PC2, color = cluster)


qmap(location = "boston", zoom = 12) + 
  geom_point(aes(x = lng, y = lat, color = cluster), data = stations)


save(file="data/stations_clustered.RData", stations, compress = TRUE)
```